\textbf{Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota}

### 2017-04-17 submission/manuscript.Rmd
updated manuscript outline

---
title: "**Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
csl: plos.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---
\vspace{35mm}

@@

- colon cancer 
- right and left tumors are morphologically and genetically distinct
- the microbiome is of healthy people and patients with CRC is distinguishable at the OTU level
- specific keystone species are associated with CRC tumors
- do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
- first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? 
- how does the microbiome vary spatially along the colon?

@@

@@

@@

### Methods

@@

>>>>

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: plos.csl
---
\vspace{35mm}

@@

- colon cancer 
- right and left tumors are morphologically and genetically distinct (why Q is important)
- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
- specific keystone species are associated with CRC tumors
- Central Question: do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations

- What data is needed to answer this question? first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa

- what methods are used to get the data? unprepped colonoscopy

- What analysis must be applied? - diversity analysis and machine learning to pull out small variations/differences among the noisy data set.

- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool

@@

- what data were obtained? 16S sequencing results for samples taken from healthy human colons that had not undergone bowel prep
Figure 1: sample collection schematic

- what were the results of the analyses? the results of the analyses are as follows:

#### Diversity analyses
- there are similar levels (within error margins) of each phyla for each location, but there is noise due to interpatient variability. Overall, there is also a similar level of diversity in each location. However, each patient has a different pattern of diversity over the locations, some are high in R, some are high in L, etc. 
Figure 2:
* 2A: relative abundance by phyla 
* 2B: simpson diversity 
* 2C: NMDS/ordination of some kind

#### similarity analyses
- we can compare sides, samples or patients by looking at distances between samples. here we used the thetayc calculator to make comparisons. The right mucosa and lumen are most similar to each other. As you move closer to the exit, the samples diverge. With the exception of the left lumen- this sample best resembles the stool collected from exit as it is... exiting.

Figure 3:
* 3A: the right mucosa and lumen are most similar to each other 
* 3B: the left lumen is most similar than exit stool

#### Random Forest classification models 
- we can use machine learning models to build classification models that identify specific features in noisy data sets that allow for separation.
- we used a Random Forest model to separate samples from each side or site

Figure 4:
* 4A: Classifying lumen vs mucosal samples on each side (L vs L, R vs R)
* 4B: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)

#### Important OTUs 
- the Random Forest model lets us pick out features that dictate the classification model separation values

Figure 5:
* 5A: There are differentially abundant OTUs in lumen vs mucosa
* 5B: There are differentially abundant OTUs in right vs left lumen, right vs left mucosa

- There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)

@@

Recap of results, interesting discussion points about diversity and previous studies of diversity along the gut axis of healthy people

Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)

-how did the analyses answer the central question?
The analyses identified OTUs that are relatively unique to each location, among a wide variation differences between study participants

-what does this answer tell us about the broader field?
This answer told us that there are specific differences of bacteria in each side and that we can use this information to form hypotheses about how specific microbial/ecological variations from the host can affect oncogenesis. It tells us too that sample collection methods are important, and what we have found in unprepped bowels do/do not match published findings.

-what's next? cancer associated OTUs on one side or the other, compare to previous studies. But really to understand what's going on in the various regions of the gut we need more information than just 16S

@@

### Methods
Study participants/IRB

Sample collection

Sample processing/DNA extraction

16S rRNA sequencing

Statistical analysis/Random Forest

@@

Figure 1

Figure 2

Figure 3

Figure 4

Figure 5

### References

### 2017-04-17 submission/manuscript.tex
updated manuscript outline

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: plos.csl
---
\vspace{35mm}

@@

Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^

@@

$\dagger$ Corresponding author: pschloss@umich.edu

1\. Department of Microbiology and Immunology, University of Michigan Medical School, Ann Arbor, Michigan 48109

2\. Pennslyvania State University, Hershey, Pennslyvania ??

3\. Department of Internal Medicine, Division of Gastroenterology, University of Michigan Medical School, Ann Arbor, Michigan

@@

### Abstract

Colorectal cancer (CRC) remains a leading cause of death worldwide. Tumors of the proximal (right) and distal (left) colon are morpologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of colorectal cancer tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. In this study, we extended our observations of the fecal microbiome to analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal biopsies, luminal and exit stool for 20 healthy individuals. Diversity analysis of each location revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the machine-learning algorithm Random Forest to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location revealed distinct populations that were found in each location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas and Actinomyces were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. Taken together, our results have identified distinct bacterial populations distinct of the proximal and distal colon. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

### Introduction

- colon cancer 
- right and left tumors are morphologically and genetically distinct (why Q is important)
- the microbiome of healthy people and patients with CRC is distinguishable at the OTU level
- specific keystone species are associated with CRC tumors
- Central Question: do the populations of bacteria in the left and right colon contribute to the development of specific subtypes of CRC? 
- nearly all studies of the gut microbiome have been from shed stool or colonoscopy with bowel prep
- bowel prep changes the mucosal/fecal microbiome and what can be analyzed from it. 
- how does the microbiome vary spatially along the colon? we know that the environment varies dramatically but we dont know how those environmental factors specifically affect bacterial populations

- What data is needed to answer this question? first, need to characterize healthy species that are specific to each location. then we can ask, if/how disruption affects tumorigenesis? so what is needed is paired samples from healthy people that have not undergone bowel prep, from the right/left and lumen/mucosa

- what methods are used to get the data? unprepped colonoscopy

- What analysis must be applied? - diversity analysis and machine learning to pull out small variations/differences among the noisy data set.

- hypothesis: there are distinct bacteria on the left and right side that can be distinguished by 16S sequencing of stool

### Results

- what data were obtained? 16S sequencing results for samples taken from healthy human colons that had not undergone bowel prep
Figure 1: sample collection schematic

- what were the results of the analyses? the results of the analyses are as follows:

#### Diversity analyses
- there are similar levels (within error margins) of each phyla for each location, but there is noise due to interpatient variability. Overall, there is also a similar level of diversity in each location. However, each patient has a different pattern of diversity over the locations, some are high in R, some are high in L, etc. 
Figure 2:
* 2A: relative abundance by phyla 
* 2B: simpson diversity 
* 2C: NMDS/ordination of some kind

#### similarity analyses
- we can compare sides, samples or patients by looking at distances between samples. here we used the thetayc calculator to make comparisons. The right mucosa and lumen are most similar to each other. As you move closer to the exit, the samples diverge. With the exception of the left lumen- this sample best resembles the stool collected from exit as it is... exiting.

Figure 3:
* 3A: the right mucosa and lumen are most similar to each other 
* 3B: the left lumen is most similar than exit stool

#### Random Forest classification models 
- we can use machine learning models to build classification models that identify specific features in noisy data sets that allow for separation.
- we used a Random Forest model to separate samples from each side or site

Figure 4:
* 4A: Classifying lumen vs mucosal samples on each side (L vs L, R vs R)
* 4B: Classifying side of lumen or side of mucosa (L vs R lumen, L vs R mucosa)

#### Important OTUs 
- the Random Forest model lets us pick out features that dictate the classification model separation values

Figure 5:
* 5A: There are differentially abundant OTUs in lumen vs mucosa
* 5B: There are differentially abundant OTUs in right vs left lumen, right vs left mucosa

- There is some commonality between the OTUs - drivers of left sidedness are consistent in both lumen and mucosa (more of this in discussion)

### Discussion

Recap of results, interesting discussion points about diversity and previous studies of diversity along the gut axis of healthy people

Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)

-how did the analyses answer the central question?
The analyses identified OTUs that are relatively unique to each location, among a wide variation differences between study participants

-what does this answer tell us about the broader field?
This answer told us that there are specific differences of bacteria in each side and that we can use this information to form hypotheses about how specific microbial/ecological variations from the host can affect oncogenesis. It tells us too that sample collection methods are important, and what we have found in unprepped bowels do/do not match published findings.

-what's next? cancer associated OTUs on one side or the other, compare to previous studies. But really to understand what's going on in the various regions of the gut we need more information than just 16S

### Acknowledgments

### Methods
Study participants/IRB

@@

### Figure Legends

Figure 1

Figure 2

Figure 3

Figure 4

Figure 5

### References

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{Machine learning identifies unique taxa differentiating proximal
and distal human colonic microbiota}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\begin{document}

\maketitle

@@

Kaitlin J. Flynn\textsuperscript{1}, Charles C.
Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2},
Danielle Kimberly Turgeon\textsuperscript{3}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

\(\dagger\) Corresponding author:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

1. Department of Microbiology and Immunology, University of Michigan
Medical School, Ann Arbor, Michigan 48109

2. Pennslyvania State University, Hershey, Pennslyvania ??

3. Department of Internal Medicine, Division of Gastroenterology,
University of Michigan Medical School, Ann Arbor, Michigan

@@

\subsubsection{Abstract}\label{abstract}

Colorectal cancer (CRC) remains a leading cause of death worldwide.
Tumors of the proximal (right) and distal (left) colon are
morpologically and genetically distinct. Previous work from our group
found that microbial dysbiosis is associated with the development of
colorectal cancer tumors in studies of both mice and humans. Analysis of
the fecal microbiota from healthy and CRC patients further revealed
different microbial signatures associated with disease. In this study,
we extended our observations of the fecal microbiome to analysis of the
proximal and distal human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure.
This technique allowed us to characterize the native proximal and distal
luminal and mucosal microbiome without prior chemical disruption. 16S
rRNA gene sequencing was performed on proximal and distal mucosal
biopsies, luminal and exit stool for 20 healthy individuals. Diversity
analysis of each location revealed that each site contained a diverse
community, and that a patient's samples were more similar to each other
than to that of other individuals. Since we could not differentiate
sites along the colon based on community structure or community
membership alone, we employed the machine-learning algorithm Random
Forest to identify key species that distinguish biogeographical sites.
Random Forest classification models were built using taxa abundance and
sample location revealed distinct populations that were found in each
location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas
and Actinomyces were most likely to be found in mucosal samples versus
luminal samples (AUC = 0.925). The classification model performed well
(AUC = 0.912) when classifying mucosal samples into proximal or distal
sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The left mucosa was found to have high
populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. Finally, comparison of all samples to fecal samples taken
at exit uncovered that the feces were most similar to samples taken from
the left lumen, again reflecting the anatomical structure of the colon.
Taken together, our results have identified distinct bacterial
populations distinct of the proximal and distal colon. Further
investigation of these bacteria may elucidate if and how these groups
contribute to differential oncogenesis processes on the respective sides
of the colon.

\subsubsection{Introduction}\label{introduction}

\begin{itemize}
\item
  colon cancer
\item
  right and left tumors are morphologically and genetically distinct
  (why Q is important)
\item
  the microbiome of healthy people and patients with CRC is
  distinguishable at the OTU level
\item
  specific keystone species are associated with CRC tumors
\item
  Central Question: do the populations of bacteria in the left and right
  colon contribute to the development of specific subtypes of CRC?
\item
  nearly all studies of the gut microbiome have been from shed stool or
  colonoscopy with bowel prep
\item
  bowel prep changes the mucosal/fecal microbiome and what can be
  analyzed from it.
\item
  how does the microbiome vary spatially along the colon? we know that
  the environment varies dramatically but we dont know how those
  environmental factors specifically affect bacterial populations
\item
  What data is needed to answer this question? first, need to
  characterize healthy species that are specific to each location. then
  we can ask, if/how disruption affects tumorigenesis? so what is needed
  is paired samples from healthy people that have not undergone bowel
  prep, from the right/left and lumen/mucosa
\item
  what methods are used to get the data? unprepped colonoscopy
\item
  What analysis must be applied? - diversity analysis and machine
  learning to pull out small variations/differences among the noisy data
  set.
\item
  hypothesis: there are distinct bacteria on the left and right side
  that can be distinguished by 16S sequencing of stool
\end{itemize}

\subsubsection{Results}\label{results}

\begin{itemize}
\item
  what data were obtained? 16S sequencing results for samples taken from
  healthy human colons that had not undergone bowel prep Figure 1:
  sample collection schematic
\item
  what were the results of the analyses? the results of the analyses are
  as follows:
\end{itemize}

\paragraph{Descriptive analysis}\label{descriptive-analysis}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  there are similar levels (within error margins) of each phyla for each
  location, but there is noise due to interpatient variability. Overall,
  there is also a similar level of diversity in each location. However,
  each patient has a different pattern of diversity over the locations,
  some are high in R, some are high in L, etc. Figure 2:
\item
  2A: relative abundance by phyla
\item
  2B: simpson diversity
\item
  2C: NMDS/ordination of some kind
\end{itemize}

\subsubsection{Discussion}\label{discussion}

-how did the analyses answer the central question?

-what does this answer tell us about the broader field?

\subsubsection{Acknowledgments}\label{acknowledgments}

\subsubsection{Methods}\label{methods}

Study participants/IRB

@@

\subsubsection{Figure Legends}\label{figure-legends}

\end{document}
